Outline

### 2023-04-25 manuscript/outline.Rmd
Adding results outline and figure notes

@@

Explain the chromatograms (RT/int/ what "goodness" requires?)

@@

@@

@@

@@

@@

Make sure labels are the same as above - probably stick with XCMS for specificity.

@@

@@

@@

@@

@@

@@

## Unstructured thoughts
Pleas:
  - Report peak boundaries! If not max/min m/z then at least max/min RT
    - Helps with visualization and raw data feature extraction
  - Run a pooled/aggregate sample

>>>>

Include the molecular mass in the feature label? If known, include name?

Put the actual good/bad label in the figure somewhere

What's the importance of the different colors?

@@

Explain the chromatograms (RT/int/ what "goodness" requires?) explain x/y axis in figure legend, maybe do in bulk across all at once?

@@

Good figure, spacing between text and labels and data

@@

Data description

  - Counts of MFs in each manually-assigned category
  - Proportion of good/bad in each dataset
  - Supplemental figure: class colored histograms with each metric?
  - Supplemental figure: class colored pairs plot?

Logistic regressions

  - Predictions are strongly binomial for all and xcms, less so for two-parameter model
    - Discuss why we dropped the p-value metric?
  - Performance within a given dataset vs across datasets and model selection
    - Introduce FDR and GPF metrics
      - Motivated by actual questions
      - FDR = 1-precision (what fraction are likely false?)
      - GPF = recall/sensitivity (what percent did I recover?)
    - Very low FDR (5-10%) and high GPF (75-90%) *within* datasets for all metrics and xcms metrics, worse for raw data (~20% FDR, ~75% GPF)
    - Higher FDR and GPF *across* datasets for all and xcms, about the same for raw data, leaves them all performing about the same(!)
    - Figure 4
    - Overfitting concerns (line slope) on cross-dataset
    - Step-down methodology and independence from training set (Figure 5)
  - Model stability
    - How similar are the model parameters to each other when trained on Falkor/MESO/both?
      - Supplemental figure: 3x errorbar plots for Falkor/MESO/both model estimates +/- SEs
    - What fraction of the dataset did we actually need to label? (Answer: basically the whole thing, possibly even more)
      - Supplemental figure: subset-based training with individual dots for min (maybe full?) model estimates added to full model estimates +/- SE bars
  - Threshold selection and FDR in CultureData/Pttime (Figure 6)
    - Higher threshold more representative of real research/applications in situations where 50% chance of being a peak is too high
    - Additional external validation for FDR that still shows increase, albeit small (0-1% -> 2-5%)
    - Unable to check GPF without fully labelling the datasets

Other ML strategies

  - Comparison to RF things
  - Comparison to glmnet things

Biological implications

  - Clockplot and univariate statistics
    - Report proportions of MFs in each category?
    - Many "bad" still have strong biology trend, not a sufficient measure on its own
    - Missing out on some good peaks (far left clockplot)
    - Do the FDR corrected p-value comparisons with more/less peaks by threshold

  - PERMANOVA table
    - Better performance somehow with 90% threshold than manually labeled "good" peaks?
    - Emphasize the lack of power when all peaks are included in the analysis with 1% and 0% thresholds
  - Actual biology things
    - Most metabolites track with biomass
    - Surface-specific largely MAAs, some new?
    - Degradation products or bacterial things accumulating at depth?

@@

Make sure the break gap is consistent

Maybe turn into two rows, one zoomed and one unzoomed?

Differentiate the 0.9 line from the axis break line - don't like the vertical black line, looks like data

Convert "meh" to "ambiguous"

@@

Is averaging the two points for the best fit line best? Lines sometimes have different slopes

Is GPF worth using instead of more common jargon like 1-precision

Maybe simpler metric

Line feels a little heavy and too dark, put behind data, maybe make line grey instead of dark black

@@

Make sure labels are the same as above - probably stick with XCMS for specificity. (Could just go in tables?)

@@

Maybe avoid "rank" phrasing? Instead of best/worst?

@@

Render as a table and see how that looks instead?

Include Pttime definition in the figure caption?

Maybe include the facet captions as the y-axis label instead? Anitra doesn't love how it looks as-is

@@

Switch the shapes to make empty circles "good peaks missed"

Add some internal circle to help render information about the overplotting in the center? Give a bar graph breakdown for the  things within the circle?

@@

Biological circumstances to test:
1. Using ALL peaks produced by XCMS
2. Using only the manually-categorized "Good" peaks
3. Using the model-categorized "Good" peaks at threshold 0.99
4. Using the model-categorized "Good" peaks at threshold 0.5
5. Using the model-categorized "Good" peaks at threshold 0.01

```{r}
library(vegan)
MS3000_metadata <- file_data %>%
  filter(samp_type=="Smp") %>%
  filter(str_detect(filename, "180821")) %>%
  select(filename, depth) %>%
  column_to_rownames("filename")
norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")

both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs <- MS3000_features %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  select(feature, feat_class, pred_prob)

wide_data <- norm_area_data %>%
  group_by(feature) %>%
  filter(var(norm_area)>0) %>%
  # mutate(norm_area=rank(norm_area)) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  left_join(pred_probs) %>% 
  pivot_wider(names_from = "filename", values_from = norm_area) %>%
  column_to_rownames("feature")

mat_data <- wide_data %>%
  # filter(feat_class=="Good") %>%
  # filter(pred_prob>0.9) %>% 
  # filter(pred_prob>0.5) %>%
  # filter(pred_prob>0.1) %>%
  select(-feat_class, -pred_prob) %>%
  data.matrix() %>%
  t()

adonis2(mat_data~depth, data = MS3000_metadata, method = "manhattan")
nmds_output <- metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0, 
                       autotransform = FALSE, noshare = FALSE, wascores = FALSE)
nmds_output$points %>%
  as.data.frame() %>%
  cbind(MS3000_metadata) %>%
  rownames_to_column("filename") %>%
  select(filename, starts_with("MDS"), depth) %>%
  mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
  ggplot(aes(x=MDS1, y=MDS2)) +
  geom_polygon(aes(fill=depth, color=depth, group=basename), alpha=0.5) +
  annotate(geom = "label", x=Inf, y=Inf, 
           label=paste("Stress =", round(nmds_output$stress, 2)),
           hjust=1, vjust=1, label.r=unit(0, "in")) +
  theme_minimal()
```

```{r}
mat_all <- wide_data
mat_good <- wide_data %>% filter(feat_class=="Good")
mat_90 <- wide_data %>% filter(pred_prob>0.9)
mat_50 <- wide_data %>% filter(pred_prob>0.5)
mat_10 <- wide_data %>% filter(pred_prob>0.1)
mat_01 <- wide_data %>% filter(pred_prob>0.01)

mat_list <- lst(mat_good, mat_90, mat_50, mat_10, mat_01, mat_all)
thresh_tests <- lapply(mat_list, function(mat_given){
  mat_data <- mat_given %>%
    select(-feat_class, -pred_prob) %>%
    data.matrix() %>%
    t()
  perm_output <- adonis2(mat_data~depth, data = MS3000_metadata, method = "manhattan")
  nmds_output <- metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0,
                         autotransform = FALSE, noshare = FALSE, wascores = FALSE)
  nmds_gp <- nmds_output$points %>%
    as.data.frame() %>%
    cbind(MS3000_metadata) %>%
    rownames_to_column("filename") %>%
    select(filename, starts_with("MDS"), depth) %>%
    mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
    ggplot(aes(x=MDS1, y=MDS2)) +
    geom_polygon(aes(fill=depth, color=depth, group=basename), alpha=0.5) +
    annotate(geom = "label", x=Inf, y=Inf, 
             label=paste("Stress =", round(nmds_output$stress, 2)),
             hjust=1, vjust=1, label.r=unit(0, "in")) +
    theme_minimal()
  list(perm_output, nmds_output, nmds_gp)
}) %>% set_names(names(mat_list))

imap(thresh_tests, function(thresh_i, thresh_name){
  as.data.frame(thresh_i[[1]][1,]) %>% mutate(thresh_name)
}) %>% bind_rows() %>%
  select(thresh_name, R2, `F`, `Pr(>F)`) %>%
  `rownames<-`(NULL) %>%
  column_to_rownames("thresh_name") %>%
  knitr::kable()

lapply(thresh_tests, `[[`, 3) %>%
  do.call(what=gridExtra::grid.arrange)
```

See which/why peaks differ between predicted and actual quality to start explaining why 90% threshold performs better than the "Good" cutoff

@@

  - Fundamental tradeoff between GPF and FDR should be determined by the scientific question
  - FDR and GPF rates are very rarely reported but critically important to include
  - Full model and XCMS model are definitely overfit
    - Steep performance drop on new data
    - Disagreement based on training set
  - Only tested on HILIC data, expectations for RP columns
  - Using medians/means as summary statistics ignores differences between treatments (i.e. lots of blanks blows up the median): maybe better to use max/top3? Maybe better to implement independently for each sample class?
  - XCMS alone does not provide enough information to distinguish good peaks from bad with much confidence

@@

## Unstructured thoughts

Journal target: BMC Bioinformatics
Alternative: PLoS Comp Bio (Too high-impact? Not as many MS papers)
Alternative: MDPI? That's where 255,000 was published
Alternative II: Bioinformatics (very bio-focused)

Pleas:
  - Report peak boundaries! If not max/min m/z then at least max/min RT
    - Helps with visualization and raw data feature extraction
  - Run a pooled/aggregate sample
